---
params:
  series: "historical-climate-data"
  resolution: "10m"
  model: NULL
  country_code: "bra"
  country_suffix: NULL
---

# LogoClim: WorldClim 2.1 Data Download

## Overview

::: {.callout-note}
This document is intended for use by the LogoClim project team. For related workflows, refer to the other notebooks in the `qmd` directory, which cover data downloading and transformation procedures.
:::

This document provides a step-by-step guide for uploading WorldClim 2.1 data to the [LogoClim OSF repository](https://osf.io/2bpjh/). For information on obtaining the data, see the [WorldClim website](https://www.worldclim.org/).

## Set the Environment

```{r}
library(beepr)
library(cli)
library(dplyr)
library(fs)
library(groomr) # github.com/danielvartan/groomr
library(here)
library(osfr)
library(prettycheck) # github.com/danielvartan/prettycheck
library(purrr)
library(readr)
library(rutils) # github.com/danielvartan/rutils
library(stringr)
library(zip) # OS independent
```

```{r}
source(here("R", "wc_license.R"))
source(here("R", "wc_readme.R"))
```

## Set the Initial Variables

### `params`

This code chunk is only used when the document is rendered from the command line. See `quarto-loop.R` in the `R` directory for more details.

```{r}
if (exists("params")) params |> list2env(envir = environment())
```

### Options

```{r}
options(cli.progress_show_after = 0)
```

### Series

Pick **one** of the following.

```{r}
#| eval: false

series <- "historical-climate-data"
```

```{r}
#| eval: false

series <- "historical-monthly-weather-data"
```

```{r}
#| eval: false

series <- "future-climate-data"
```

### Resolution

Pick **one** of the following.

```{r}
#| eval: false

resolution <- "10m"
```

```{r}
#| eval: false

resolution <- "5m"
```

```{r}
#| eval: false

resolution <- "2.5m"
```

```{r}
#| eval: false

resolution <- "30s"
```

```{r}
#| eval: false

resolution <- "all"
```

### Directories

Check the paths to the directories where the data will be stored.

```{r}
raw_data_dir <- here("data-raw")
```

```{r}
raw_data_wc_dir <- path(raw_data_dir, "worldclim")
```

```{r}
raw_data_wc_series_dir <- path(raw_data_wc_dir, series)
```

```{r}
raw_data_wc_series_res_dir <- path(
  raw_data_wc_dir,
  series,
  resolution |> str_replace_all("\\.", "\\-")
)
```

```{r}
data_dir <- here("data")
```

```{r}
data_series_dir <- path(data_dir, series)
```

```{r}
dirs <- c(
  raw_data_dir, raw_data_wc_dir, raw_data_wc_series_dir,
  raw_data_wc_series_res_dir, data_dir, data_series_dir
)

for (i in dirs) {
  if (!dir_exists(i)) {
    dir_create(i, recurse = TRUE)
  }
}
```

## Zip the Files

```{r}
#| eval: false

model <- NULL # "ACCESS-CM2"
country_code <- "bra"
country_suffix <- NULL # "box" # "mainland"
```

```{r}
#| eval: true

if (length(series) == 1) {
  switch(
    series,
    "historical-climate-data" = {
      series_suffix <- "hcd"
    },
    "historical-monthly-weather-data" = {
      series_suffix <- "hmwd"
    },
    "future-climate-data" = {
      series_suffix <- "fcd"
    }
  )
} else {
  series_suffix <- NULL
}
```

```{r}
#| eval: false

zip_file <-
  data_dir |>
  path(
      paste0(
      country_code,
      ifelse(!is.null(country_suffix), paste0("-", country_suffix), ""),
      paste0("-", str_replace_all(resolution, "\\.", "\\-")),
      ifelse(!is.null(series_suffix), paste0("-", series_suffix), ""),
      ifelse(!is.null(model), paste0("-", tolower(model)), ""),
      ".zip"
    )
  )

zip_file
```

```{r}
#| eval: false

zip_file |>
  zip::zip(
    files =
      data_dir |>
      dir_ls(
        type = "file",
        regexp = "\\.zip$|\\.gitignore$",
        recurse = TRUE,
        invert = TRUE
      ) |>
      path_rel(data_dir),
    root = data_dir
  )

beep(1)
```

```{r}
#| eval: false

osf_files <- zip_file
```

### Add LICENSE and README Files

```{r}
#| eval: false

osf_files <-
  data_dir |>
    path(c("LICENSE.md", "README.md")) |>
    c(zip_file) |>
    unname()

osf_files
```

## Upload the Files to OSF

```{r}
#| eval: false

assert_internet()

osf_auth()
```

```{r}
#| eval: false

if (resolution == "10m") {
  osf_id <- "x7fcm"
} else if (resolution == "5m") {
  osf_id <- "un5jm"
} else if (resolution == "2.5m") {
  osf_id <- "4vxet"
} else if (resolution == "30s") {
  osf_id <- "qgyha"
}

osf_id
```

```{r}
#| eval: false

osf_id |>
  osf_retrieve_node() |>
  osf_upload(
    path = osf_files,
    conflicts = "overwrite",
    progress = TRUE,
    verbose = TRUE
  )

beep(1)
```

## Other Resources (Optional)

### Rename Data Directory Files

```{r}
#| eval: false

files <- data_dir |> dir_ls()

files |>
  file_move(
    path(
      data_dir,
      files |> basename() |> str_to_lower()
    )
  )
```

### Rename Files in OSF

```{r}
#| eval: false

assert_internet()

osf_auth()
```

```{r}
#| eval: false

if (resolution == "10m") {
  osf_id <- "x7fcm"
} else if (resolution == "5m") {
  osf_id <- "un5jm"
} else if (resolution == "2.5m") {
  osf_id <- "4vxet"
} else if (resolution == "30s") {
  osf_id <- "qgyha"
}

osf_id
```

```{r}
#| eval: false

osf_files <-
  osf_id |>
  osf_retrieve_node() |>
  osf_ls_files(
    pattern = "5m",
    n_max = Inf
  )

osf_files |> print(n = Inf)
```

```{r}
#| eval: false

osf_files <-
  osf_files |>
  dplyr::mutate(
    name =
      name |>
      str_to_lower() |>
      str_replace_all("\\.", "\\-")
  )

osf_files |> print(n = Inf)
```

### Delete Files in OSF

::: {.callout-caution}
Careful! There is no coming back from this.
:::

```{r}
#| eval: false

assert_internet()

osf_auth()
```

```{r}
#| eval: false

if (resolution == "10m") {
  osf_id <- "x7fcm"
} else if (resolution == "5m") {
  osf_id <- "un5jm"
} else if (resolution == "2.5m") {
  osf_id <- "4vxet"
} else if (resolution == "30s") {
  osf_id <- "qgyha"
}

osf_id
```

```{r}
#| eval: false

osf_files <-
  osf_id |>
  osf_retrieve_node() |>
  osf_ls_files(
    pattern = "_2-5m",
    n_max = Inf
  )

osf_files
```

```{r}
#| eval: false

# for (i in split(osf_files, seq_len(nrow(osf_files)))) {
#   i |> osf_rm(check = FALSE)
# }
```
